```{r}
#install.packages("dplyr")
library(dplyr)

library(ggplot2)
```

# Practicing `dplyr` functions using the `starwars` dataset

```{r}
head(starwars)
```

## Subset rows with `filter`

`filter()` allows us to subset the rows of a dataset based on the values

```{r}
# filter for rows with human characters
filter(starwars, species == "Human")

# save to a new R object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

# filter multiple columns at once
filter(starwars, species == "Human", sex == "male")
```

-   `!` is "not:

-   `==` is "equal"

-   `&` is "and"

-   `|` is "or"

-   `<` is "less than"

-   `>` is "greater than"

-   `is.na()` returns TRUE if values is NA

```{r}
# return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
# get characters that are NOT human
filter(starwars, species != "Human")
```

variable_name `%in%` vector_of_values returns TRUE if the value of variable_name is equal to one of the values in vector_of_values (example below)

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") #returns false befause "grey" is not in vector of colors

# filter for characters with brown or blond hair
filter(starwars, hair_color %in% c("brown","blond"))
```

`grepl("characters_of_interest", variable_name)` returns TRUE if the value of variable_name contains the characters_of_interest

```{r}
filter(starwars, grepl("blond", hair_color))
# "blonde" and "blond"

# get all characters with blue in their eyes
filter(starwars, grepl("blue", eye_color))
```

You can use `>`, `<`, `<=`, `>=` to compare numeric or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

    -   name \< "d"

        -   a, b, c, d

-   ordinal categorical variables (factors) are ranked based on the built-in-order

```{r}
# starwars characters with a name starting with letters before "D"
filter(starwars, name < "D")
```

```{r}
str(diamonds)
# cut has 5 built-in levels

filter(diamonds, cut > "Good")

"D" < "d" # false
"d" < "D" # true
```

`between(x, left, right)` is a shorthand for x \>= left & x \<= right

```{r}
between(5,1,7) # true because 5 is >= 7 and <= 1
filter(starwars, between(mass, 50, 100))
```

### Practice

1.  missing hair color

```{r}
filter(starwars, is.na(hair_color))
```

1.  shorter than 100cm and not droids

```{r}
filter(starwars, height < 100, species != "Droid")
```

1.  born between years 100 and 200

```{r}
filter(starwars, between(birth_year,100,200))
```

1.  another of choice

```{r}
# str(starwars)
filter(starwars, height > 100, hair_color == "blond", sex == "male", species == "Human")
```

```{r}
filter(starwars, mass > 50, hair_color == "white", between(height, 100, 200))
```

```{r}
filter(starwars, homeworld == "Serenno", species == "Human")
```

## Order rows with `arrange`

`arrange()` reorder rows based on the values of column(s)

```{r}
arrange(starwars, name)
# order of character values: aAbBcCdD

# order by species and birth year
arrange(starwars, species, birth_year)
# NA rows get pushed to bottom

# use desc() to go in descending order
arrange(starwars, species, desc(birth_year))
```

### Practice

1.  arrange from tallest to shortest

```{r}
arrange(starwars, desc(height))
```

1.  arrange from shortest to tallest by species

```{r}
arrange(starwars, height, species) # order by height, tiebreaker is alphabetic species
arrange(starwars, species, height) # orer by species, within species, order shortest to tallest
```

1.  alphabetize characters by their hair color then eye color

```{r}
arrange(starwars, hair_color, eye_color)
```

1.  one of choice

```{r}
arrange(starwars, homeworld, films)
```

## Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
# name to eye color column (inclusive)
select(starwars, name:eye_color) 
```

We can also remove columns by name using a `-`

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

## Helpful functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything())
# put species first, then all other columns

# alphabetize columns
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

-   `starts_with("abc")` matches column names that start with "abc"

-   `ends_with("abc")` matches column names that end with "abc"

-   `contains("abc")` matches column names that contain "abc"

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```

### Practice

1.  How many humans live in X homeworld?
2.  6 ways:
    1.  select(starwars, hair_color, skin_color, eye_color, birth_year)
    2.  select(starwars, hair_color:birth_year)
    3.  select(starwars, -name, -height, -mass, -sex, -gender, -homeworld, -species, -films, -vehicles, -starships)
    4.  select(starwars, -c(name, height, mass, sex, gender, homeworld, species, films, vehicles, starships)))
    5.  select(starwars, -c(name:mass,sex:starships))
    6.  
select(starwars, contains("\_"))
    7.  select(starwars, contains("color"), birth_year))
    8.  select(starwars, ends_with("color"), ends_with("year"))

```{r}
names(starwars)
```

```{r}

select(starwars, ends_with("color"), ends_with("year"))
```

## Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

We can also create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

transmute(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: `sum(), mean()`

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small,
       raw_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

Group by a variable using the `.by` parameter

```{r}
mutate(starwars_small,
      species_mass = mean(mass, na.rm = T),
       .by = species)
```

Change the type of a variable

```{r}
mutate(starwars_small,
       mass = as.character((mass)))

# for 1/0 columns can use as.logical to change to TRUE/FALSE
```

### Practice

1.  convert mass to pounds (1 kg \~ 2.2 lbs)

```{r}
mutate(starwars_small,
      mass_lbs = mass*2.2)
```

## Summarize data with `summarize`

```{r}
summarize(starwars, 
          mean_mass = mean(mass, na.rm = T))

# group by species
summarize(starwars, 
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

The only columns kept are the column you group by and the columns you create.

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars, 
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)

# group by multiple groups
summarize(starwars, 
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = c(species,sex))
```
