---
---
---

```{r}
# loading in library
library(dplyr)
```

-   filtering the observations of your dataset based on one or more variable using filter()

```{r}
head(data)

filter(data, age > 70)
```

-   create a smaller dataset with a subset of variables using select()

```{r}
names(data)

# subset of data with all columns measuring level of X in blood
sub_data = select(data, TSH, T3, TT4, T4U, FTI, TBG)
sub_data
```

-   Add 2 new columns to your dataset using mutate()

```{r}
filter(sub_data, between(FTI, 53, 142))

# 1 if their level is normal, 0 if it is not
# 52-142 is normal for FTI, 13-39 is normal for TBG
mutate(sub_data, 
       normal_FTI = case_when(FTI > 53 & FTI < 142 ~ 1, 
                              FTI < 53 | FTI > 142 ~ 0),
       normal_TBG = case_when(TBG > 13 & TBG < 39 ~ 1, 
                              TBG < 13 | TBG > 39 ~ 0))
```

-   Create a data table of grouped summaries on at least one numeric variable by at least one categorical variable using summarize()

```{r}
# average TBG level based on sex
summarize(data, 
          mean_TBG = mean(TBG, na.rm = T),
          .by = sex)
```

-   Reorder a data table of your choice by one or more variables using arrange()

```{r}
# arranging based on age and target value
arrange(data, age, target)
```

-   Create at least one new visualization using some form of an updated dataset

```{r}
# adding column to sub_data
sub_data <- mutate(sub_data, normal_FTI = 
       case_when(FTI > 53 & FTI < 142 ~ 1, 
                  FTI < 53 | FTI > 142 ~ 0))

# bar plot visualization of boolean var vs. double var
ggplot(data = sub_data, aes(x = normal_FTI, y = FTI)) +
  geom_bar(stat = "summary",
           fun = "mean",
           fill = "lightblue")

# ~ 150 people with not normal FTI levels and ~100 with normal levels
```
